A performance assessment of evolutionary algorithms in a volunteer computing system

### 2016-03-12 PPSN2016/paper.Rnw
Adds table

\title{Evolutionary algorithms in a volunteer computing system:
  performance assessment}

@@

In general, a distributed volunteer-based evolutionary computation
system based on the browser is simply a client-server system
whose client is, or it can be, embedded in the browser via
JavaScript. Since JavaScript is  the only language that is present
across all browsers, the choice was quite clear. We should emphasize
that NodIO is more intended as an auxiliary computing engine, more
than the main one, so performance of JavaScript as a language is not
so important; even so, we have made a comparison between JavaScript and other languages
\cite{2015arXiv151101088M} that shows that performance of JavaScript
is comparable to other interpreted languages; compiled languages would
be faster, but, of course, it is impossible to gather volunteers
spontaneously and without any installation with them.

In this sense, in this paper we propose the {\sf NodIO} framework, a
cloud or bare metal based volunteer evolutionary computing system
derived from the {\sf NodEO} library, whose architecture has been
developed using JavaScript on the client as well as the server.
All parts of the framework are free and available with a free license
from \url{https://github.com/JJ/splash-volunteer}.

Thus, {\sf NodIO} architecture has two tiers:\begin{enumerate}
\item A REST server, that is, a server that includes several {\em
  routes} 
  that can be called for storing and retrieving information (the ''CRUD'' cycle:
  create, request, update, and delete) from the server. 
  A JSON data format is used for the communication between 
  clients and the server. There are two kinds of information:
  {\em problem} based, that is, related to the
  evolutionary algorithm such as {\tt PUT}ing a chromosome in or {\tt
  GET}ing a random chromosome from it, and {\em information} related
  to the performance and state of the experiments. It also performs logging
  duties, but they are basically a very lightweight and high performance
  data storage \cite{jj:idc:lowcost}.
  The server has the capability to
  run a single experiment, storing the chromosomes in a data structure
  that is reset when the solution is found. This data structure can
  hold every chromosome in a particular experiment, or have a finite
  size that erases the oldest chromosomes once it has filled to
  capacity. In this paper we will test both implementations.  
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays graphs, some
  additional links, and information on the experiment. This code runs
  an evolutionary algorithm {\em island} that starts with a random
  population, then after every 100 generations, it sends the best individual
  back to the server (via a {\tt PUT} request), and then requests a random
  individual back from the server (via a {\tt GET} request). We have
  kept the number of generations between migrations fixed since it is
  a way of finding out how many real work every client has done. 
\end{enumerate}

Figure \ref{fig:system} describes the general system architecture and
algorithm behavior. Different web technologies, such as JQuery or {\tt
  Chart.js} have been used to build the user interface elements of the
framework, a part of which is shown in Figure \ref{fig:screenshot} and
should be running in \url{http://nodio-jmerelo.rhcloud.com}.

\begin{figure}[!t]
\centering
\includegraphics[width=0.95\linewidth]{nodio-screenshot.png}
\caption{Screenshot showing the fitness and number of users as seen by
volunteers. The one on the left shows fitness, on the right the
accumulated number of users. \label{fig:screenshot}}
\end{figure}
%
\begin{figure}[!t]
\centering
\includegraphics[width=3in]{system.pdf}
\caption{Description of the proposed system. Clients execute a JavaScript EA
  in the browser, which, every 100 generations, sends the best
  individual and receives a random one back from the server.}
\label{fig:system}
\end{figure}

JavaScript is a functional language, so in order to work with a
problem, a fitness function must be supplied at the creation of the algorithm object, 
called {\tt Classic}. In this case the classical Trap 
function \cite{Ackley1987} has been used. Depending on the
problem additional handing functions and configurations 
can be supplied.

The next Section will describe experiments performed to
establish a baseline performance and gather initial performance
results. In the first set of experiments, performed last year, we used the 40-trap
problem, while the current experiments changed to the more difficult 50-trap
in order to compare the performance with a more computationally intensive problem.

%---------------------------------------------------------------
\section{Modeling the performance of a volunteer-based distributed computer} 
\label{sec:experiments}

\begin{table}[htb]
\caption{Experiment table, with summary of results. \label{tab:runs}}
\begin{center}
\begin{tabular}{l|rrl}
\hline
Experiment & \#Runs & Different IPs & Traps \\
\hline
April 4th 4/4 & 57 & 191 & 40 \\
April 24th 4/24 &  231 & 559& 40  \\
July 31th 7/31 & 97 & 179 & 40 \\
\hline
February, cache=128 & 61 & 75 & 50  \\
February, cache=64 & 61 & 220 & 50  \\
February, cache=32 & 39 & 86 & 50  \\
% grep start ~/Code/splash-volunteer/data/2016/nodio-2016-2-28-cache=32.log
\hline
\end{tabular}
\end{center}
\end{table}
%
Initial experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to \$0.00. Experiments were
announced through a series of posts on Twitter and, in the latest case, Telegram, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the month of February. All  %% April --> January ??
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:runs}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
%Pablo: but it would great at least describe how many times was announced and how long
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 50-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions resets to the void set. There was no special intention to wait
until all clients had finished, thus it might happen that. In fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is indeed negligible. %Pablo: cite

>>>>

\title{A performance assessment of evolutionary algorithms in a volunteer computing system}

@@

In general, a distributed volunteer-based evolutionary computation
system based on the browser is simply a client-server system
with a client embedded in the browser via JavaScript. We call this
system {\sf NodIO}. All parts of the framework are free and available
with a free license from
\url{https://github.com/JJ/splash-volunteer}.

The architecture of the {\sf NodIO} system is divided in two tiers:\begin{enumerate}
\item A REST (representational state transfer) server that responds to
  HTTP requests made by the browsers. These HTTP requests use JSON for
  encoding information, in this case chromosomes and associates
  information send by the clients and return requests made by the
  server.  The server has the capability to
  run a single experiment, storing the incoming chromosomes in a data structure
  that is reset when the solution is found. This data has a finite
  size that erases the oldest chromosomes once it has filled to
  capacity.  
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays the fitness and
  other graphs, some
  additional links, and information on the experiment. This code runs
  an evolutionary algorithm {\em island} that starts with a random
  population, then after every 100 generations, it sends the best individual
  back to the server (via a {\tt PUT} request), and then requests a random
  individual back from the server (via a {\tt GET} request). We have
  kept the number of generations between migrations fixed since it is
  a way of finding out how much real work every client is doing. 
\end{enumerate}

Figure \ref{fig:system} describes the general system architecture and
algorithm behavior. Different web technologies, such as JQuery or {\tt
  Chart.js} have been used to build the user interface elements of the
framework, which should be running in
\url{http://nodio-jmerelo.rhcloud.com}, a free resource hosted in the
OpenShift Platform as a Service.\footnote{It is not guaranteed to be running,
or running the same version, when you read this, however; you can always get
the sources from GitHub and set it up yourself.}
%
\begin{figure}[!t]
\centering
\includegraphics[width=3in]{system.pdf}
\caption{Description of the proposed system. Clients execute a JavaScript EA
  in the browser, which, every 100 generations, sends the best
  individual and receives a random one back from the server.}
\label{fig:system}
\end{figure}

NodIO needs a fitness function to work with; in this case the classical Trap 
function \cite{Ackley1987} has been used; this function runs in the
browser, but the server needs to know when the solution has been
found, so a termination condition is also included as a configuration
option in the server. In all cases we are using 50 Traps, a problem
that is difficult enough to need the intervention of several users for
an extended amount of minutes.

All experiments are announced in the same way: we use social networks,
mainly Twitter, to announce the start of a new experiment. LinkedIn,
Facebook, and also private groups in Telegram and WhatsApp have been
used too for announcements. We also encourage friends and followers to
put an original post for announcing it, instead of just sharing, in
the way allowed by the platform, the original post. During the running
of the experiments, users are engaged, questions answered, and all
kind of explanations given if requested. In general, announcements
have been made in Spanish, but in some cases English was used
too. This makes for a certainly unrealiable experimentation framework,
but it is a realistic one. The steps in the design of the evolutionary
algorithm will be explained next.

%---------------------------------------------------------------
\section{Experiments and results} 
\label{sec:experiments}

\begin{table}[htb]
\caption{Experiment table, with summary of results. \label{tab:runs}}
\begin{center}

<<experiments, cache=FALSE,echo=FALSE>>=
experiment.aggregates$Experiment <- c("base","graph1","graph2","graph3","reboot","cachereboot")
kable(experiment.aggregates)
@ %def

\end{center}
\end{table}
%
Initial experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to \$0.00. Experiments were
announced through a series of posts on Twitter and, in the latest case, Telegram, and
results were published here \cite{DBLP:conf/gecco/GuervosG15}. For the
purpose of this paper, we repeated the announcement several times
through the month of February. All  %% April --> January ??
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:runs}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
%Pablo: but it would great at least describe how many times was announced and how long
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 50-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions resets to the void set. There was no special intention to wait
until all clients had finished, thus it might happen that. In fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is indeed negligible. %Pablo: cite
